HO unification from object language to meta language

### 2024-02-23 paper/main.tex
update tex

\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$ and we assume that the unification of our meta language is correct:
%
$$
t_1 \Ue t_2 \mapsto \sigma \Rightarrow \sigma t_1 \Ee \sigma t_2
$$

@@

We write $s \in \mathcal{L}_\lambda$ if all unif variables in
$s$ are applied to distinct bound variables.

\begin{align}
t_1 \Uo t_2 \mapsto \rho &\Rightarrow \rho t_1 \Eo \rho t_2\\
s_i \in \mathcal{L}_\lambda \land \exists \rho, \rho s_1 \Eo \rho s_2 &\Leftrightarrow s_1 \Uo s_2 \mapsto \rho' \subseteq \rho \\
\forall \rho, \rho s_1 \Eo \rho s_2 &\Rightarrow \rho t_1 \Uo \rho t_2
\end{align}

@@

\input{lang_descr.tex}

@@

A key property needed in unification is being able to verify if two terms are
equal. This is kind of a structural equality verification between two terms,
where variable dereferencing is performed when the variable is assigned. A
sketch of the equality function is given for the OL language in
\cref{code:term-equal}. Though, this equality relation over terms of a language
can be powered by other reduction rules depending equational theory being
considered. In our case, the OL terms are equal under $\eta\beta$ redex. This mean
that new rules for those two redexes are added in the implementation of
\elpiIn{fo_equal}.

>>>>

\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
\newcommand{\llambda}{\ensuremath{\mathcal{L}_\lambda}\xspace}

@@

\newcommand{\specunif}[3]{
  #3_i \in \mathcal{L}_\lambda \Rightarrow %
    \exists \rho, %
      \rho #3_1 #1 \rho #3_2  %
        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho 
}

\newcommand{\unifcorrect}[3]{
  % \forall \rho #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #3_i \in \llambda \Rightarrow
      #3_1 #2 #3_2 \mapsto \rho
        \Rightarrow 
          \rho #3_1 #1 \rho #3_2  %
}

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #3_i \in \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho 
}

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$ and we assume that the unification of our meta language is correct:

%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

@@

We write $s \in \mathcal{L}_\lambda$ if all unif variables in
$s$ are applied to distinct bound variables.
%
\begin{gather}
\unifcorrect{\Eo}{\Uo}{s}\\
\unifcomplete{\Eo}{\Uo}{s}\\
% \forall \rho'\rho, 
  \rho s_1 \Eo \rho s_2 \Rightarrow
  \rho' \subseteq \rho \Rightarrow
  \rho's_i \in \mathcal{L}_\lambda \Rightarrow
  \rho' s_1 \Uo \rho' s_2
\end{gather}

@@

% \input{lang_descr.tex}

@@

A key property needed in unification is being able to verify if two terms are
equal wrt a given equational theory. This relation allow to compare terms under
a certain substitution mapping, so that any time a variable $v$ is assigned in a
subterm, a dereferencing of $v$ is performed. After variable dereferencing, the
test for equality is continued on the new-created subterm.

The base equality function over terms can be defined as follows:

\begin{figure}
  \begin{elpicode}
    % congruence
    type ~(\Eo)~ tm -> tm -> o.
    app L1 ~\Eo~app L2 :- forall2 ~(\Eo)~ L1 L2.
    lam F1 ~\Eo~lam F2 :- pi x\ x ~\Eo~x => F1 x ~\Eo~F2 x.
    con X ~\Eo~con X.
    fov N ~\Eo~fov N.
    % eta
    lam F ~\Eo~T :- not (T = lam _), 
      pi x\ beta T [x] (T' x), lam F ~\Eo~lam T'.
    T ~\Eo~(lam F) :- not (T = lam _), 
      pi x\ beta T [x] (T' x), lam T' ~\Eo~lam F.
    % beta
    app [lam X | TL] ~\Eo~T :- beta (lam X) TL T', T' ~\Eo~T.
    T ~\Eo~app [lam X | TL] :- beta (lam X) TL T', T ~\Eo~T'.
  \end{elpicode}
  \caption{Equal predicate}
  \label{code:eq-pred}
  \Description[code:equal-pred]{code:equal-pred}
\end{figure}

Though, this equality relation is to be extended with $\eta\beta$ reduction
rules to mirror the theory of the OL. This mean
that new rules for those two redexes are added in the implementation of
\elpiIn{fo_equal}.
